A Type System for String Sanitation\\Implemented Inside a Python

### 2014-04-16 sanitation-plas14/sanitation-plas14.tex
PLAS version used in today's meeting with JA.

\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter

@@

@@

@@

@@

\begin{thm}
  Let $S$ be a term in the source language. If $\sctx \vdash S : \psi$ 
  then $S \sreduces S'$ and $S' : \psi$.
\end{thm}

\begin{thm}
  Let $P$ be a term in the target language. If $\sctx \vdash P : \theta$ 
  then $P \sreduces P'$ and $P' : \theta$.
\end{thm}

@@

\begin{thm}
If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
\begin{itemize}
\item (a) $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
\item (b) $T \err$ and $S \err$.
\end{itemize}
\end{thm}
\begin{proof}
By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
\begin{itemize}
\item Tr-string. By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
By source language type safety, it follows that $s \in \lang{r}$.
\item Tr-Concat.
By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
By induction on the left premise, $T_1 \treduces \tstr{s_1}$, $S_1 \sreduces \str{s_1}$ and $s_1 \in \lang{r_1}$
or else $T_1 err$. 
Similarly for $T_2$ and $S_2$.
In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\siconcat{S_1}{S_2} \err$.
In the non-error case, $\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}$ by T-E-Concat.
Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
\item T-subst.
Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
By inverison, $S_1 : \strin{r_1}$ and $S_2 : \strin{r_2}$.
Note that $s : \subst{r}{S_1}{S_2} : \stringin{\tsubst{r}{r_1}{r_2}}$ by S-T-Subst.
By induction, $\trden{S_1} = T_1$ and $\trden{S_2} = T_2$.
Either (a) or (b) holds for each.

@@

If (b) holds for either the left or right premise, then $\sisubst{r}{S_1}{S_2} \err$ by the propagation rules corresponding to S-T-subst, 
and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.
\end{itemize}
\end{proof}

\subsection{Type Safety Proof}

\subsection{New Theorems}

\begin{thm}[Progress for Internal Language]
\end{thm}

\begin{thm}[Preservation for Internal Language]
\end{thm}

\begin{thm}[Representational Consistency]
If $\cdot \vdash e : \tau \sreduces i$ then $\cdot \vdash i : \sigma$ and $\sigma \equiv \trep(\tau)$.
\end{thm}

\begin{thm}[Main Soundness Result]
  If $\cdot \vdash e : \strin{r} \sreduces i$ then either $i \treduces \err$ or
  if $i \sreduces v$ for $v$??? then $v \in r$.
\end{thm}

%\begin{thm}[Preservation]
%  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
%  For all terms $e$:
%  \begin{itemize}
%    \item If $e:T$ and $e:T \treduces i$ then $i : \sigma$ such that $h(T) = \sigma$.
%    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
%  \end{itemize}
%\end{thm}
%\begin{proof}
%The proof is a straightforward induction on the derivation of the combined evaluation relation.
%\begin{itemize}[label=$ $,itemsep=1ex]
%  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
%  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
%  \item \textbf{E-String}. By the definition of typing for internal terms, $\tstrf{e} : istr$. It suffices to show that $h(\str) = istr$, which follows from the definition of $h$.
%  \item \textbf{E-String\_in}. By the definition of typing for internal terms, $\tstrf{e} : istr$. It suffices to show that $h(\strin[r]) = istr$, which follows from the definition of $h$ for arbitrary $r$.
%  \item \textbf{E-concatval}. By the definition of typing for internal terms, $\tstrf{e_1} + \tstrf{e_2} : istr$. So it suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
%  \item \textbf{E-concatR, E-concatL}. Consider E-concatR. By induction, $e_1' : \strin{l_1}$. By inversion of T-Concat at the premise, $e_2 : \strin{r_2}$. Therefore, $e_1 + e_2 : \strin[r_1 + r_2]$. The left rule is symmetric.
%  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
%    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
%    By the Coercion Theorem, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
%    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
%  \item \textbf{E-Filter}. By inversion, $e : \strin[r']$ $\reduces e'$ so by the induction $e':\strin[r']$.
%    Therefore, \\${\tt filter}(\strin[r], e'):$ $\strin[r \backslash r' + \emptyset]$ by T-Filter.
%  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = istr$, which is true by definition.
%  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
%    It follows that $[\strin[r']](e') : \strin[r']$.
%  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
%  \item \textbf{E-DConvertval}. By the definition of typing for internal terms, $\ifilter(r,\tstrf{e}) : istr$. It suffices to show that $h(\strin[r_1+r_2]) = istr$, which follows from the definition of $h$.
%\end{itemize}
%\end{proof}
%
%\begin{thm}[Progress]
%  If $e:T$ then $e:T \reduces^* e' \treduces^* i$ where $i \ival$.
%\end{thm}
%\begin{proof}
%By induction on the derivation of $e:T$.
%For \textbf{T-Equiv-Include}, note that $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval. 
%We have that $e \val$ and $e : \strin[r]$, so $e \treduces \tstrf{e}$ by E-string\_in.
%The remaining cases follow by induction on the hypotheses and application of corresponding evaluation rules.
%\end{proof}
%
\section{Implementation Inside a Python}
%
%TODO-nrf rewrite
%The $\lcs$ language is implemented as a type system extension in Ace; this extension
%is illustrated in the examples at the beginning of this paper.
%
%Computing a regular expression representating the language
%$R \backslash S$ is necessary in order to type-check expressions in which the filter function
%occurs. This language is computed by translating $R$ and $S$ into finite automata,
%complementing the final states of $S$, then constructing the cross-product of $R$
%and $S'$. Type checking terminates only because this construction is decidable.
%
%In addition to this construction, other more typical operations -- such
%as equality checks for regular expressions and regular expression matching -- are
%also necessary. For these reasons, the Ace extension implementing $\lcs$ includes a library
%implementing each of these constructions.
\newcommand{\F}[1]{\textsf{#1}~}
\newcommand{\FF}[1]{\textsf{#1}}
\newcommand{\Q}{\FF{Arg}}
\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lamAce}}
\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\texttt{ace}}$}
\begin{figure*}
\begin{tabular}{ l r l l }
$\tau$ & & $\xlA{\tau}$ & \xA{\tau}\\
\hline
$\strin{r}$ & & $\fvar{stringin}[r]$ & \verb|string_in["|$r$\verb|"]|\\
  $e$ & & $\xlA{e}$ & \xA{e}\\
  \hline
  ${\tt str}[r, s]$ & \text{synthetic position} & $\FF{intro}[\FF{str}[s]]()$ : \fvar{stringin}[\FF{rx}[$r$]] & \verb|"|$s$\verb|" (string_in["|$r$\verb|"])|\\
  & \text{analytic position} & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
  ${\tt concat}(e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{concat}](\xlA{e_2})$ & \xA{e_1}\verb| + |\xA{e_2} \\
  ${\tt replace}[r](e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{replace}~\FF{rx}[r]](\xlA{e_2})$ & \xA{e_1}\verb|.replace("|$r$\verb|", |\xA{e_2}\verb|)|\\
  ${\tt coerce}[r](e)$ & & $\xlA{e}\cdot\FF{elim}[\tvar{coerce}~\FF{rx}[r]]()$ & \xA{e}\verb|.coerce("|$r$\verb|")|
\end{tabular}
\caption{Translation of $\lcs$ to $\lamAce$ and Ace.}
\end{figure*}
\begin{figure}
\small\begin{flalign}
& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
	\tlam{tyidx}{\FF{R}}{
	\tlam{a}{\klist{\Q}}{\\
& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
	}}
}\}\\
& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
	\tlam{a}{\klist{\Q}}{\\
& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
}}\\
& \quad\quad}{opidx'}{d}}
}\}
\}
\end{flalign}
\caption{Definition of ...}
\end{figure}
\begin{figure}
\[
\begin{array}{lcl}
% & & Definition & Kind\\
\tvar{concat} & := & \tinl{\kunit, R+R}{b}\\%& \kunit + (R + R)\\
\tvar{replace} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinl{R,R}{\tvar{r}}}}\\% & \karrow{a}{b}\\
\tvar{coerce} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinr{R,R}{\tvar{r}}}}\\
\end{array}
\]
\caption{Definitions}
\end{figure}

>>>>

\documentclass[10pt,preprint]{sigplanconf}
\newcommand{\lamAce}{\lambda_{\text{Ace}}}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{afterpage}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{verbatimbox}
\input{../att-icfp14/macros-atlam}
\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{light-gray}{gray}{0.5}
\usepackage{listings}
\usepackage{wasysym}
    \makeatletter
\usepackage{enumitem}
\usepackage{enumerate}

@@

\afterpage{

@@

\clearpage
} % End \afterpage

@@

\begin{lem}
  If $\sctx \vdash S : \stringin[r]$ then $r$ is a well-formed regular expression. 
\end{lem}
\begin{proof}
  By induction on the typing derivation and the closure lemmas for regular
  expressions.
\end{proof}

@@

\begin{thm}
  Let $S$ be a term in the source language. If $\sctx \vdash S : \psi$ 
  then $S \sreduces S'$ and $S' : \psi$.
\end{thm}
\begin{proof}
  TODO-nrf

By a straightforward induction on the derivation of $\sctx \vdash S : \psi$.
Note that preservation depends upon the substitution correspondence theorem
for the T-T-Subst case and the closure lemmas for the T-T-Concat case.
\end{proof}

\begin{thm}
  Let $P$ be a term in the target language. If $\tctx \vdash P : \theta$ 
  then $P \sreduces P'$ and $P' : \theta$.
\end{thm}
\begin{proof}
TODO-nrf

  By a straightforward induction on the derivation of $\tctx \vdash P : \theta$.
Note that the well-formedness of regular expressions indexing ${\tt rx}$ types
depends upon the closure lemmas.
\end{proof}

@@

\begin{thm}
If $S : \strin{r}$ then there exists a T such that $\trden{s} = T$ and either: 
\begin{description}
  \item[(a)] $T \treduces \tstr{s}$ and $S \sreduces \strin{s}$, and $s \in lang{r}$.
  \item[(b)] $T \err$ and $S \err$.
  \end{description}
\end{thm}
\begin{proof}
By induction on the derivation of $\trden{S} = T$. TODO-nrf add case magic.
\begin{description}
\item[Tr-string] By assumption $\trden{S} = \tstr{s}$. The (a) property follows.
By S-E-Str, $S \sreduces \str{s}$ and by T-E-Str, $\tstr{s} \treduces \tstr{s}$.
By source language type safety, it follows that $s \in \lang{r}$.
\item[Tr-Concat]
By inversion of the typing relaiton for the source lnaguage, $S_1 : \strin{r_1}$ and $S_2 : \stringin{S_2}$.
By induction on the left premise, $T_1 \treduces \tstr{s_1}$, $S_1 \sreduces \str{s_1}$ and $s_1 \in \lang{r_1}$
or else $T_1 err$. 
Similarly for $T_2$ and $S_2$.
In the error cases, the propagations rules corresponding to S-E-Concat and T-E-Concat imply that $\tconcat{T_1}{T_2} \err$ and $\siconcat{S_1}{S_2} \err$.
In the non-error case, $\siconcat{S_1}{S_2} \sreduces \strin{s_1 \cdot s_2}$ by S-E-Concat and $\tconcat{T_1}{T_2} \treduces \tstr{s_1 \cdot s_2}$ by T-E-Concat.
Finally, $s_1 \in \lang{r_1}$ and $s_2 \in \lang{r_2}$ implies that $s_1 \cdot s_2 \in \lang{r}$ by \ref{lemmaregex}.
\item[Tr-subst]
Suppose $\sisubst{r}{S_1}{S_2} : \strin{s}$ for some $x$. 
By inverison, $S_1 : \strin{r_1}$ and $S_2 : \strin{r_2}$.
Note that $s : \subst{r}{S_1}{S_2} : \stringin{\tsubst{r}{r_1}{r_2}}$ by S-T-Subst.
By induction, $\trden{S_1} = T_1$ and $\trden{S_2} = T_2$.
Either (a) or (b) holds for each.

@@

If (b) holds for either the left or right premise, then $\sisubst{r}{S_1}{S_2} \err$ by the propagation rules corresponding to S-T-subst, 
and $\tsubst{S_1}{S_2} \err$ by the propagation rules corresponding to T-T-Subst.

\item[Tr-Coerce-Ok] TODO-nrf.

\item[Tr-Coerce-NotOk] TODO-nrf.

\end{description}
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% END SECTION TWO -- THIS IS THE CONFLICT LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\F}[1]{\textsf{#1}~}
\newcommand{\FF}[1]{\textsf{#1}}
\newcommand{\Q}{\FF{Arg}}
\newcommand{\xlA}[1]{\lfloor #1 \rfloor_{\lamAce}}
\newcommand{\xA}[1]{$\lfloor #1 \rfloor_{\texttt{ace}}$}
\begin{figure*}
\begin{tabular}{ l r l l }
$\tau$ & & $\xlA{\tau}$ & \xA{\tau}\\
\hline
$\strin{r}$ & & $\fvar{stringin}[r]$ & \verb|string_in["|$r$\verb|"]|\\
\\
  $e$ & & $\xlA{e}$ & \xA{e}\\
  \hline
  ${\tt str}[r, s]$ & \text{synthetic position} & $\FF{intro}[\FF{str}[s]]()$ : \fvar{stringin}[\FF{rx}[$r$]] & \verb|"|$s$\verb|" (string_in["|$r$\verb|"])|\\
  & \text{analytic position} & $\FF{intro}[\FF{str}[s]]()$ & \verb|"|$s$\verb|"|\\
  ${\tt concat}(e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{concat}](\xlA{e_2})$ & \xA{e_1}\verb| + |\xA{e_2} \\
  ${\tt replace}[r](e_1; e_2)$ & & $\xlA{e_1}\cdot\FF{elim}[\tvar{replace}~\FF{rx}[r]](\xlA{e_2})$ & \xA{e_1}\verb|.replace("|$r$\verb|", |\xA{e_2}\verb|)|\\
  ${\tt coerce}[r](e)$ & & $\xlA{e}\cdot\FF{elim}[\tvar{coerce}~\FF{rx}[r]]()$ & \xA{e}\verb|.coerce("|$r$\verb|")|
\end{tabular}
\caption{Translation of $\lcs$ to $\lamAce$ and Ace.}
\end{figure*}
\begin{figure}
\small\begin{flalign}
& \F{tycon}\fvar{stringin}~\F{of}\FF{R}~\{\\
& \quad \F{iana}\{\tlam{opidx}{\FF{String}}{
	\tlam{tyidx}{\FF{R}}{
	\tlam{a}{\klist{\Q}}{\\
& \quad\quad \tvar{arity0}~\tvar{a}~(\tvar{check}~\tvar{opidx}~\tvar{tyidx}~\titerm{\iup{\tvar{opidx}}})
	}}
}\}\\
& \quad \F{esyn}\{\tlam{opidx}{\kunit+(\FF{R}+\FF{R})}{
	\tlam{a}{\klist{\Q}}{\\
& \quad\quad \tsumcase{\tvar{opidx}}{\_}{\\
& \quad\quad\quad \tvar{arity2}~\tvar{a}~\tlam{a1}{\Q}{\tlam{a2}{\Q}{\\
& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a1}~\tlam{r1}{\FF{R}}{\tlam{i1}{\kITerm}{~}}\\
& \quad\quad\quad\quad \tvar{rsyn}~\tvar{a2}~\tlam{r2}{\FF{R}}{\tlam{i2}{\kITerm}{~}}\\
& \quad\quad\quad\quad\quad (\ttype{stringin}{\FF{rseq}(\tvar{r1}; \tvar{r2})}, \\
& \quad\quad\quad\quad\quad\titerm{{\tt iconcat}(\iup{\tvar{i1}}; \iup{\tvar{i2}})})
}}\\
& \quad\quad}{opidx'}{d}}
}\}
\}
\end{flalign}
\caption{Definition of ...}
\end{figure}
\begin{figure}
\[
\begin{array}{lcl}
% & & Definition & Kind\\
\tvar{concat} & := & \tinl{\kunit, R+R}{b}\\%& \kunit + (R + R)\\
\tvar{replace} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinl{R,R}{\tvar{r}}}}\\% & \karrow{a}{b}\\
\tvar{coerce} & := & \tlam{r}{R}{\tinr{\kunit, R+R}{\tinr{R,R}{\tvar{r}}}}\\
\end{array}
\]
\caption{Definitions}
\end{figure}
